---
title: "Project_2_C"
author: "Suyog Dharmadhikari & Magaritte Nguyen"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
params:
  api-key: "a2ac9584c657495ab1447aeef32b97d3"
---

<!-- 
############################################################# 
#############################################################
Project 2 - Group 2C
############################################################# 
############################################################# 
-->
## Spoonacular Recipe and Food API Vignette


<!-- link to Wiki page on APIs: -->
<!-- [API] (https://en.wikipedia.org/wiki/API) -->

This document is a vignette to show how to retrieve data from an [API](https://spoonacular.com/food-api/docs). To demonstrate, we will be interacting with the Spoonacular Recipe and Food API. We are going to to build a few functions to interact with some of the endpoints and explore some of the data that we can retrieve for the Spoonacular API.

We will use a function that connects with `complexSearch`, which is an end point of the API search. It returns a list of lists containing recipe names and other variables of interest that have default values or can be user defined. The user may either enter the type of cuisine, recipes with a specific ingredient, number of recipes to output based on modifications provided, type of diet, and protein range. We have another function that can retrieve the data for multiple cuisines specified by the user as well. We have a third function which can also be used to contact other endpoint of the API, `findByNutrients` to get all the recipes based on carbohydrate content.
 
## Requirements

To use the functions for interacting with the Spoonacular Recipe and Food API, we used the following packages:

- [`tidyverse`](https://www.tidyverse.org/): tons of useful features for data manipulation and visualization
- [`jsonlite`](https://cran.r-project.org/web/packages/jsonlite/): API interaction

In addition to those packages, I used the following packages in the rest of the document:

- [`httr`](https://cran.r-project.org/web/packages/httr/): extra functionality for 
- [`knitr`](https://cran.r-project.org/web/packages/knitr/index.html): displaying tables in a markdown friendly way

**Note**: User needs to go to [API_link](https://spoonacular.com/food-api/console#Profile), create a profile and  generate a **key**. This key will be used to `authenticate` the user for each API calls.

<!-- setup -->
```{r setup, echo = FALSE, eval = TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)  
library(httr)
library(jsonlite)
```

# API Interaction Functions

For this section, we have written some functions that will connect the user to the Spoonacular API and returns a object with the required information specified by the user. 

<!-- Magaritte Nguyen's API Key -->
<!-- 1faaf57a0fbc433fa74f31a9419dc85b  -->

## EndPoint - 1

### `my_Spoonacular_API`

In this function we call `complexSearch` endpoint of the Spoonacular API. The user can use this function to get the data from the API for different types of cuisine, recipes with a specific ingredient, maximum number of recipes to output, type of diet, and protein range. The function is user friendly and handles upper and lower case scenarios with spaces. **`GET`** from `httr` is used to connect with the API and retrieved the required information while `fromJSON` is used to convert the raw data into a readable format.

Just some extra information on how to use the `my_spoonacular_api`: we are only able to specify one cuisine at a time, or else a message might appear letting the user know what is appropriate in this field. We can also specify one ingredient, but if there is nothing that matches, then the results will just be 0 recipes.if the user would like some nutrient information as well, they can input TRUE or FALSE for this parameter. Lastly, we have different diets that the user can specify, such as vegetarian etc., and the amount of protein in grams (min and max range).

**NOTE**: user cannot fetch more than 100 records in the single call as it is restricted by the API.
```{r apiEndpointsCall, echo=TRUE, eval=TRUE}

# Query will only return words in the title of the recipe
# If set to true, you get nutritional information about each recipes returned.
# 100 is the max number of results you can get back for a single query.
my_Spoonacular_API <- function (cuisinetxt ="American", ingredient = "fruit", nutrientinfo = "TRUE", number=100, diet = "", minProtein = 10, maxProtein = 100 ){

  #valid cuisines to enter into the cuisine parameter
  #using tolower() to make everything lowercase for east of search
  #using str_replace_all for searches that are more than one word e.g. Latin American
  if ( ! any(str_replace_all(tolower(cuisinetxt), " ", "-") %in% str_replace_all( tolower(c("African", "American", "British", "Cajun", "Caribbean", "Chinese", "Eastern European", "European", "French", "German", "Greek", "Indian", "Irish", "Italian",   "Japanese", "Jewish", "Korean", "Latin American", "Mediterranean", "Mexican", "Middle Eastern", "Nordic", "Southern", "Spanish", "Thai", "Vietnamese")), " ", "-" ) ) ) { stop ('Please only include an acceptable cuisine input seen below: 

"African", "American", "British", "Cajun", "Caribbean", "Chinese", "Eastern European", "European", "French", "German", "Greek", "Indian", "Irish", "Italian",   "Japanese", "Jewish", "Korean", "Latin American", "Mediterranean", "Mexican", "Middle Eastern", "Nordic", "Southern", "Spanish", "Thai", "Vietnamese" 

*** If you include more than one cuisine seperated by a comma, ignore this message.') 
  } #end of cuisine check
  
  #using GET to return info about a topic that we are interested in 
  myGET <- GET(url=paste0("https://api.spoonacular.com/recipes/complexSearch?cuisine=", cuisinetxt, "&includeIngredients=", ingredient, "&addRecipeNutrition=", nutrientinfo, "&number=",number,"&diet=",diet,"&minProtein",minProtein,"&maxProtein",maxProtein,"&apiKey=",params$`api-key`))  
  
  #looking at content (raw data) and converting it into something we can read from myGET  
  #data
  spoonacular <- fromJSON(rawToChar(myGET$content))
  spoonacular$cuisinetxt <- cuisinetxt
  
  #results returned as a dataframe
  return(spoonacular)
  
} #end of function that called API

```

### `my_multiCuisine_API`

In this function we call `complexSearch` endpoint of the API. It is similar to `my_Spoonacular_API` but in this function the user can enter multiple cuisines to get the combined data of all the cuisines. The user has to use `paste` with `collapse = ","` as the parameter to create a list of cuisines before passing it to the function.

**NOTE**: user cannot fetch more than 100 records in the single call as it is restricted by the API.
```{r multiCusine, echo = TRUE, eval = TRUE, warning = FALSE, message=FALSE}
# Function for multiple cuisines in the cuisinetxt parameter
  my_multiCuisine_API <- function (cuisinetxt =" ", ingredient = "", nutrientinfo = "TRUE", number=100, diet = "", minProtein = 10, maxProtein = 100 ){
    
   myGETTemp <- GET(url=paste0("https://api.spoonacular.com/recipes/complexSearch?cuisine=", cuisinetxt, "&includeIngredients=", ingredient, "&addRecipeNutrition=", nutrientinfo, "&number=", number,"&diet=",diet,"&minProtein",minProtein,"&maxProtein",maxProtein,"&apiKey=",params$`api-key`))  
  
   return(myGETTemp)
  }
   
```

## EndPoint - 2

### `findRecipeByNutrients`

The `findByNutrients` endpoint of the API is called in the method below. The function can assist users in retrieving data on the nutritional value of recipes. To choose dishes from the specified **carbohydrate content** and certain number of recipes output, the user can alter the carbohydrates range. It returns the object along with some information on the recipe, its nutrient content, and other parameters. **`GET`** from `httr` is used to connect with the API and get the required information while `fromJSON` is used to convert the raw data into readable format.

**NOTE**: user cannot fetch more than 100 records in the single call as it is restricted by the API.
```{r secondAPIendpointCall, echo=TRUE,eval=TRUE,warning=FALSE,message=FALSE}

# Looking for the given number of recipes with minCarbs and maxCarbs
findRecipeByNutrients <- function(minCarbs = 10, maxCarbs = 100, numberOfRecipes = 100){
   getURL <- paste0("https://api.spoonacular.com/recipes/findByNutrients?minCarbs=",minCarbs,"&maxCarbs=",maxCarbs,"&number=",numberOfRecipes,"&apiKey=",params$`api-key`)
  foodData <- GET(getURL)
  findByNutrientsDF <- fromJSON(rawToChar(foodData$content))
  return(findByNutrientsDF)   
}

```

In the code chunk below, we have called `my_Spoonacular_API()` and `findRecipeByNutrients()`, which creates objects for different cuisines with different ingredients and other parameters. The user can reference this section of the vignette as an example call, if they want to get information about different cuisines with different modifications.
```{r CallapiEndpointSection, echo=TRUE, eval=TRUE}

#Example calls for my_Spoonacular_API
AmericanCuisineInfo <- my_Spoonacular_API (cuisinetxt ="American", ingredient = "fruit", nutrientinfo = "TRUE", number=100, diet = "", minProtein = 0, maxProtein = 100)

LatinAmericanCuisineInfo <- my_Spoonacular_API (cuisinetxt = "Latin American", ingredient = "salt", nutrientinfo = "TRUE", number=100, diet = "", minProtein = 10, maxProtein = 100)

ChineseCuisineInfo <- my_Spoonacular_API (cuisinetxt = "Chinese", ingredient = "cabbage", nutrientinfo = "TRUE", number=100, diet = "glutenFree", minProtein = 10, maxProtein = 100)

#Example calls for findRecipeByNutrients
dispDF <- findRecipeByNutrients(numberOfRecipes = 20)

```
These are the data frames containing nutrient information about the particular cuisine, which can be used further for our data exploration (EDA).
```{r viewDataFrame, echo=TRUE, eval=TRUE}

  AmericanCuisineInfotbl <- as_tibble(AmericanCuisineInfo$results)
  AmericanCuisineInfotbl
  LatinAmericanCuisineInfoTbl <- as_tibble(LatinAmericanCuisineInfo$results)
  LatinAmericanCuisineInfoTbl
  ChineseCuisineInfoTbl <- as_tibble(ChineseCuisineInfo$results)
  ChineseCuisineInfoTbl
  dispDFTbl <- as_tibble(dispDF) 
  dispDFTbl
```
# Data Exploration

## Adding New Variables and Contingency Table

In this section we have created two new variables and added them to the data frame and also created three contingency tables explaining the **association** between the variables of the dataframes. The nutrients tibble is extracted from the `AmericanCuisineInfo` data frame.

### Adding New Variables
In the below code chunk the information extracted from the API does not have information related to **total price** of the recipe and **proteins per serving**. Some variables of interest provided in the data was information about `price per serving`, `total servings` and `total_protein`. We used these variables and created two new variables `totalprice` and `protein_per_serving`. 

It was interesting to find out that total price provided in the API was in pennies. So, for `total price` we calculated this by multiplying `price per serving` and `servings` then dividing by 100 to get a dollar amount for `total price`. For `protein per serving` we divided amount of protein in a recipe by total number of servings the recipe made. 

Some nutrients that we were interested in were calories, protein, fat, and carbohydrates.
```{r varAdd, echo=TRUE, eval=TRUE}
# Example for data exploration
# American cuisine info object that was created above
CuisineInfo <- AmericanCuisineInfo

# viewing nutrients data frame nested inside results list
View(CuisineInfo$results$nutrition$nutrients[[1]])

# want to look at calories, protein, carbs, fat
# bind_rows - binding row wise for all lists
# nutrients is a list of dataframes
nutrients <- bind_rows(CuisineInfo$results$nutrition$nutrients, .id="df") %>%
  as_tibble() %>% 
  filter (name %in% c("Calories", "Protein", "Fat", "Carbohydrates")) %>%
  pivot_wider(id_cols=df, names_from=c(name, unit), values_from=amount) %>% 
  bind_cols (CuisineInfo$results[,c("id","title", "vegetarian", "vegan", "pricePerServing", "readyInMinutes", "servings")]) %>%
  mutate (totalprice = (pricePerServing*servings)/100, protein_per_serving = (Protein_g/servings ) )
# results is a data frame that we can use to do EDA
# cost per serving is in pennies.
# I converted total price/cost to make the recipe in USD

# output nutrients tibble
nutrients
```

### Contingency Table
In the below chunk of code, we created contingency tables to find the associations between `vegan and vegetarian`, `Source and Healthy`, `Source and Expensive`.

In the `vegan and vegetarian` contingency table, we see that there are **5 vegan recipes** in the **American** Cuisine with fruit being the ingredient that is being searched. So, we can conclude that some vegetarian recipes are vegan as well. This is as expected.

In the `Source and Healthy` contingency table, we could see that most of the unhealthy food recipes are from **Foodista** source. Now we know which website to avoid for healthy food options. 

In the `Source and Expensive` contingency table, we added one more variable to the dataframe called **expensive**, which was a category we defined if the price per serving was higher than Q3. Here, we can see that the recipes suggested by **Foodista** has the **maximum number** of **expensive** recipes, but also suggests **maximum number** of **inexpensive** recipes.
```{r contTable, echo=TRUE, eval=TRUE}

# make 3 contingency tables
# vegetarian, vegan 
VegConTbl <- base::table("Vegan" = nutrients$vegan, "Vegetarian" = nutrients$vegetarian)
VegConTbl

# sourceName and Healthy
sourceHealthyConTbl <- base::table("Source" = CuisineInfo$results$sourceName, "Healthy" = CuisineInfo$results$veryHealthy ) 
sourceHealthyConTbl

#create category
#price expensive
#price in pennies
new_CuisineQuery <- CuisineInfo$results %>% 
  mutate ( "expensive" = pricePerServing >= quantile(CuisineInfo$results$pricePerServing, 0.75))

sourceExpensivecontTable <- base::table("Source" = CuisineInfo$results$sourceName, "Expensive" = new_CuisineQuery$expensive ) 
sourceExpensivecontTable
```

## Summarizing The Data

In this section we are summarizing the data by using the following functions `mean()`, `median()`, `desc()`, `select()`, `summarise()`, and `sppaly()`.

In the following code chunk, for the `mean_protein` object, we did a mean price breakdown with a `group_by()` function with vegetarian and non vegetarian. In the `highProteinRecipes` object, we can see that **Cinnamon French Toast Sticks** has higher protein than **Turkey Burgers with Slaw** even though it is a vegetarian recipe. This is surprising, but also makes sense due to the protein content in eggs.
```{r summData, echo=TRUE, eval=TRUE}
# mean price breakdown by group - vegetarian and non vegetarian
# from data above in other code chunk
mean_protein <- nutrients %>%
  group_by (vegetarian) %>%
  summarise("Count" = n(), "Mean" = mean(Protein_g), "SD" = sd(Protein_g), "Med." = median(Protein_g), "Min."=min(Protein_g), "Max." = max(Protein_g))


# protein content per serving size (for the recipes)
# view the data to see what the protein per serving is.
highProteinRecipes <- nutrients %>% 
  select (vegetarian, vegan, title, protein_per_serving, Protein_g, servings) %>%
  arrange (desc(protein_per_serving)) %>%
  head(5) 

```
In the code chunk below we have created objects by fetching the information of multiple cuisines from the Spoonacular API. `paste()` is used along with `collapse` to pass as a parameter to the `my_multiCuisine_API()` custom function. The code chunk returns a tibble containing the number of recipes for each cuisine in the data that we extracted for all the ingredients and given protein range. It can be seen that the number of **American** cuisine recipes are much **higher** in comparison to **Chinese** and **Latin American** cuisines.
```{r summData2, echo=TRUE, eval=TRUE}
  
  # Vector of multiple cuisines
  cuisinetxt <- c("American", "chinese", "Latin American")
  
  # Using paste to create a text string of cuisine types to fetch information 
  cuisineComb <- paste(cuisinetxt, collapse = ",")
  
  # Calling my_multiCuisine_API function to get the combined information
  MultiCuisineInfoObj <- my_multiCuisine_API(cuisinetxt = cuisineComb, ingredient = "", nutrientinfo =  
                  "TRUE", number=100, diet = "", minProtein = 10, maxProtein = 100)    
  
  # Object with parsed data collected from API   
  spoonacularTemp <- fromJSON(rawToChar(MultiCuisineInfoObj$content))
  
  # Summarizing n counts based on cuisines 
  MultiCuisineInfo <- spoonacularTemp$results %>%
    group_by(cuisines) %>% 
    summarize(count=n()) %>% 
    filter(cuisines != "character(0)")
   
  # unlist the list of cuisine names and combining it into single name using apply
  cuisineInfoTemp <- apply(MultiCuisineInfo, 1, function(y) sapply(y, function(x) paste(unlist(x), 
                           collapse="_")))
   
  # Taking transpose of the df to makes it useful for plotting
  cuisineInfoTempTranspose <- t(cuisineInfoTemp)
   
  # Converting it to tibble
  MultiCuisineInfoTbl <- as_tibble(cuisineInfoTempTranspose)
   
  # The count variable was character, used sapply to conver it into numeric
  MultiCuisineInfoTbl$count <- sapply(MultiCuisineInfoTbl$count, as.numeric)
  
  # Multicuisine output tibble 
  MultiCuisineInfoTbl
   
```

## Plots
In this section, we did data exploration through multiple plots - like **bar plot**, **histogram**, **boxplot**, **scatter plot**, and **scatter plot with linear model**. 

In the code chunk below, we have used a bar plot to show the recipe count variations among the different cuisines. In this code chunk we have used `geom_bar()` for bar plot and `theme()` to tilt the x-axis names as they are overlapping. It can be seen from the plot that **American** cuisine has **highest** number of recipes with all the ingredients with protein range between 10-100.
```{r barPlot, echo=TRUE, eval=TRUE}

barPlot <- function(df, xVar, yVar){
  
   g <- ggplot(data = df , aes(x = xVar, y = yVar))
   g + geom_col(aes(fill = as.factor(xVar))) + scale_fill_discrete(name = "Type of Cuisine") +
   xlab("Cuisines") + ylab("Count") +
   ggtitle("Barplot For Cuisines Type Counts") + theme(axis.text.x = element_text(angle = 18))
}

barPlot(MultiCuisineInfoTbl,MultiCuisineInfoTbl$cuisines,MultiCuisineInfoTbl$count)
   
```

In this histogram, we have plotted for **total price variation** for **American** cuisine. We have used `geom_histogram()` for the histogram plot and added another layer to it using `geom_vline()`, showing the mean of **total prices** on the histogram plot. The histogram plots does not look like a normal distribution. It looks like it skewed towards the right. Maybe in future data exploration, we can transform the data into something more familiar and be able to draw conclusions from that familiar distribution.
```{r histPlot, echo=TRUE, eval=TRUE}

histPlot <- function(df, xVar, bin_width = 2){
  g <- ggplot(data = df , aes(x = xVar))
  g + geom_histogram(binwidth = bin_width, color="blue", fill="white") +
  geom_vline(aes(xintercept=mean(xVar)),
            color="red", linetype="dashed", size=1) + 
  xlab("total prices") + ggtitle("Histogram for totalprices")
}

histPlot(nutrients, nutrients$totalprice)

```
In this box plot we are looking to determine how much protein is present in the vegetarian and non vegetarian meals. We have used `geom_boxplot()` for this scatter plot and added a jitter layer using `geom_point()`. When looking at the protein content in the vegetarian and non vegetarian recipes, protein is not non-existent in vegetarian meals. This is a common misconception. The number of recipes for each category also do not differ much. Also, we see that the protein content in non vegetarian meals in generally much greater than the protein content in vegetarian meals. Even the outliers are only a little more than the median protein content in the non vegetarian meals. Something to consider is how many serving size each recipe is. This may contribute to a large difference in protein. 
```{r boxPlot, echo=TRUE, eval=TRUE}
# boxplot()
# setting the stage to add layers
# how much protein is in vegetarian meals 
g <- ggplot(nutrients, aes(x = vegetarian, y = Protein_g))
# boxplot for Protein by Vegetarian 
g + geom_boxplot() + 
    labs(title = paste0("Boxplot for Protein (g) vs. Vegetarian Recipes (", CuisineInfo$cuisinetxt, " Cuisine)"), x = "Vegetarian Recipe", y = "Protein (g)") +
    geom_point(aes(color=vegetarian), position="jitter") 
```
In this box plot we will be looking at how much protein is in vegetarian meals per serving. We have used `geom_boxplot()` for this scatter plot and added a jitter layer using `geom_point()`. From the plot, interestingly enough, when broken down by serving, the protein content in vegetarian meals (for some) have a higher protein content then the non vegetarian meals. Compared to the boxplot above this is an interesting finding.
```{r boxPlot1, echo=TRUE, eval=TRUE}
# boxplot()
# setting the stage to add layers
# how much protein is in vegetarian meals per serving to see if there is any difference...
g <- ggplot(nutrients, aes(x = vegetarian, y = protein_per_serving))
# boxplot for Protein by Vegetarian 
g + geom_boxplot() + 
    labs(title = paste0("Boxplot for Protein (g) per Serving vs. Vegetarian Recipes (", CuisineInfo$cuisinetxt, " Cuisine)"), x = "Vegetarian Recipe", y = "Protein (g) / Serving") +
    geom_point(aes(color=vegetarian), position="jitter")

```
In this scatter plot, we tried to see the relationship between protein and carbohydrates of a particular cuisine. Here, `geom_point()` was used and we can see that there is **no real pattern** to be observed between carbohydrates and protein. A few points may indicate increase when either one is increased, but over all protein stays low no matter how the carbohydrate content varies.
```{r scatterPlot, echo=TRUE, eval=TRUE}
# seperate scatter plot
# Carbohydrates_g vs. Protein_g
# setting the stage to add layers
g <- ggplot(nutrients, aes(x = Carbohydrates_g, y = Protein_g))
# scatter plot with linear regression line for carbs by protein
g + geom_point(aes(color=Protein_g), position="jitter") +
    labs(title = "Scatterplot for Carbohydrates (g) vs. Protein (g)", x="Carbohydrates (g)", y="Protein (g)") 

```
In this scatter plot, we have added a layer for the regression line using the `geom_smooth()` function. After seeing the plot, we can see that the regression line basically has **No change** in protein (g) at any amount of carbohydrates. This is similar to the plot above.
```{r scatterLMPlot, echo=TRUE, eval=TRUE, message=FALSE,warning=FALSE}
#setting the stage to add layers
g <- ggplot(nutrients, aes(x = Carbohydrates_g, y = Protein_g))
#scatter plot with linear regression line for carbs by protein
g + geom_point(aes(color=Protein_g), position="jitter") +
    labs(title = "Carbohydrates (g) vs. Protein (g)", x="Carbohydrates (g)", y="Protein (g)") + 
    geom_smooth(method = lm) 
```
# Wrap up

<!-- Need to write the summary in brief -->
To summarize everything that we did in this vignette please see the following: 

- Built custom functions to interact with some of the Spoonacular Food and Recipe API endpoints like ComplexSearch and findByNutrients

- Retrieved some of the data based on different user modifications

- Did exploratory data analysis (EDA):

  - Using contingency tables
  
  - Numerical summaries like mean, median, min, max, standard standard deviaiton
  
  - Data visualization methods like plotting -- bar plot, box plot, histogram, scatter plot, scatter plots with linear models overlayed. 

Some interesting findings for our specific vingette examples:

- Some vegetarian recipes are vegan as well

- Most of the unhealthy food recipes are from the **Foodista** source

- Recipes suggested by **Foodista** has the **maximum number** of **expensive** recipes, but also suggests **maximum number** of **inexpensive** recipes

- **Cinnamon French Toast Sticks** has a higher protein content than a non vegetarian meals like the **Turkey Burgers with Slaw**, even though it is a vegetarian recipe. This is surprising, but also makes sense due to the high protein content in eggs

- It can be seen that the number of **American** cuisine recipes are much **higher** in comparison to **Chinese** and **Latin American** cuisines for our vignette for the recipes containing fruit.

- The protein content in non vegetarian meals in generally much greater than the protein content in vegetarian meals as expected

- When broken down by serving, the protein content in vegetarian meals (for some) have a higher protein content then the non vegetarian meals

- There is **no real pattern** to be observed between carbohydrates and protein

- The regression line basically has **No change** in protein (g) at any amount of carbohydrates, confirming out plot that is only a scattet plot between protein and carbohydrates

We hope this vignette will help the user to interact with spoonacular API and do the required exploratory analysis.


